Three Problems for Sophistication

### 2015-01-15 Facticity/Facticity/paper.tex
intro v1

\title{Four Problems with Sophistication}

@@

In this paper we describe four fundamental problems that arise with all existing proposals, showing each of them to be \emph{unsound}. Based on these results we we lay down the view that this unsoundness is fundamental: while many nontrivial and useful structural properties of binary strings can be detected without a priori assumptions, it is \emph{not} possible to objectively quantify its sophistication.
\end{abstract}

\section*{Steven's playground}

Requirements for sophistication:

(1) Being an information measure, the value should count the number of bits required for an effective description of the meaningful properties of a binary object.

(2) The Invariance Theorem for Kolmogorov complexity states that a change in reference UTM can only influence the complexity up to an additive constant term. Analogously, if sophistication is to be interpreted as an objective property of the data, then there should be strict limits on the degree to which a change in the UTM can affect the sophistication.

(3) Sophistication should not be $\gamma$-equivalent to 0. While a valid information measure, the function 0 is clearly at odds with the intuition that there is no universal cap on the amount of meaningful information in the world.

(4) Sophistication should not be $\gamma$-equivalent to the Kolmgorov complexity, because then there is no point in defining a new word for the same concept. Moreover, Kolmogorov complexity is at odds with the intuition that there is no meaningful information in a random string.

@@

\noindent Kolmogorov complexity is an objective measure of the information content of an object. Although its definition contains several arbitrary choices, these only affect its value up to a constant additive term. This property is crucial for the measure's success: it allows us to view Kolmogorov complexity as a \emph{property of the data} rather than an arbitrary function computed on it. This, in turn, allows us to define the intrinsic entropy, or randomness, inherent in a single object expressed as a binary string, without any assumptions on the process that produced it. However, this notion of complexity does not in all respects match our intuitive idea of which objects are complex, and which objects are not.

Imagine an old-fashioned, analog television set which can be tuned to three frequencies. The first produces a blank screen, the second an ordinary TV show, and the third white noise. Judging by their Kolmogorov complexity, these  broadcasts have increasing information content. The first is easiest to describe, the second is harder but still contains a great deal of regularity, and the third cannot be compressed. However, intuitively speaking the second signal clearly conveys more meaningful information than the other two. This notion has proved far harder to capture in formal terms than the simple notion of `raw' information content.

Over the years one solution in particular has been proposed in many different forms. We select a description for the data by two-part coding; describing a model first, and then the data in terms of the model. The size in bits, used to describe the model is taken as a measure of the useful information in our string. This idea has been proposed many times in many different guises, which shows that it is both a persistent intuition but also that it is a difficult idea to formalize in a meaningful way. For the purposes of this paper, we will refer to this concept as \emph{sophistication}\footnotemark.

\footnotetext{As the historical overview shows, sophistication is only one name for this idea, and not the first. The alternatives, however, are more awkward, and sophistication has been used by different authors for different realizations of the same principle.}

The intuitive idea of sophistication makes one crucial demand: we can find an objectively value for the model complexity of a string, either by having a single objectively optimal two-part description, or a set of descriptions with equivalent model complexity. In the framework of Kolmogorov complexity, two observers cannot disagree by more than a constant about the information content of a dataset. In formal terms: the Kolmogorov complexity of a dataset $x$ is defined as the shortest effective description for $x$ in some Turing complete description language. This description language is usually expressed as a universal Turing machine $U(\bar\imath y) = \phi_i(y)$ where $\phi_i$ is the $i$'th partial computable function in some effective enumeration, and the $\bar\imath$ is a self-delimiting encoding of $i$. If two people disagree about the shortest effective description of $x$, we can interpret this as them using different description languages, ie. different universal Turing machines. The crucial aspect of Kolmogorov complexity, the starting theorem that makes it interesting as a function at all, is \emph{invariance}: two people using different description languages will get the same value for a string's Kolmogorov complexity, albeit up to a fixed constant. Without invariance, Kolmogorov complexity would be an arbitrary function. With invariance, we can be sure that we are talking about an intrinsic property of the data. The same consideration must be given to each proposal for formalising sophistication: is it invariant?

There are many approaches to sophistication, but the following pattern is common to all of them. We start with a space of \emph{models}. There can be finite sets, computable probability distributions or simple functions from strings to strings. The function perspective is the most generic, all other model classes can be translated to it. The approach to sophistication is to find a representation for a dataset $x$ in two parts: a computable function $f$ and and input $y$, such that $f(y) = x$. As with Kolmogorov complexity, the shorter a description is, the better. We take the shortest description and those close to it, and consider them \emph{candidates}. How close a description must be to the optimum differs between variants, one popular choice is to use a pre-chosen constant value. We then choose a single description from the candidate set and return the description length of the model as the sophistication. Usually, the candidate with the smallest model determines the sophistication.

The implicit assumption behind sophistication is that this approach will separate the description of the string into a structural part, representing its regularities, and a stochastic part representing its non-deterministic properties. In other words, the string is cast as a random choice from a model for which it is `typical'.

In this article we will show that most proposals are not invariant. \sdr{Ik wil dit nog steeds graag sterker gesteld hebben.} For one remaining case, we can offer no such proof, but no proof of invariance exists either. This is likely due to the complex nature of the model class chosen, and the difficulty of resolving the matter of invariance shows how complex such a sophistication measure would be to build on, even if it did prove to be invariant.

\subsection{A brief historical overview}

We first provide a brief chronological overview of the existing variants of sophistication. We will return to each in-depth in the body of the paper.

The first suggestion that a representation of a string can be separated cleanly into an encoding of its regular and its stochastic proeperties came from Kolmogorov himself, in a talk given at a conference in Tallinn in 1973 (first committed to paper by Cover \cite{cover1985kolmogorov} and later discussed more extensively discussed by Vereshchagin and Vit\'anyi \cite{vereshchagin2004kolmogorov}). Here, he proposed what has become known as the \emph{Kolmogorov structure function}.Minimizes over two part representations for a given string $x$, consisting of some finite set $S$ (containing $x$) and the index of $x$ in $S$. The \emph{minimal sufficient statistic} is the set with the lowest complexity for which the length of the two-part representation reaches the prefix-free Kolmogorov complexity. It was first shown by Shen \cite{shen1983concept} that there are strings for which the best model cannot be small.

Koppel \cite{koppelSoph1988,koppel1991almost} built on this notion, coining the phrase \emph{sophistication}. He changed the class of models from finite sets to total functions and used monotone functions, instead of prefix-free Kolmogorov complexity. We expect that this change was made to extend the theory to infinite strings, an approach which has been abandoned by subsequent authors. We will only consider the finite version (based on regular functions) in this paper.

While Koppel expands the model class from finite sets to functions,he does require that the functions be total. This is a recurring theme: if the model class is expanded to all computable functions, the universal Turing machine $U$ itself becomes a model. If we assume that $U$ computes the function $\phi_u$ in the standard enumeration of computable functions, and that $\bar\imath y$ is a shortest description for $x$ on $U$ (ie. $U(\bar\imath y) = x)$), then $\br{u}\br{\imath}y$ is also a description for $x$ on $\bar u$. This description has a constant model size ($|\br{u}|$), ie. every $x$ has constant sophistication. This `collapse' to a universal model is the reason that most treatments of sophistication constrain the model class to exclude the universal Turing machine. This makes it more expensive to move information out of the model. As we shall see, defining sophistication quickly becomes a balancing act, forcing information one way or the other.

In 1996 Gell-Mann and Lloyd introduced `Effective complexity' \cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (largely independently, it seems) at a notion that fits the mold of sophistication. Effective complexity describes a binary string in terms of a probability distribution over finite strings (an `ensemble' in the terminology of Gell-Mann and Lloyd) and the entropy of the distribution. The reasoning is that, if the string is typical for the distribution, the entropy is sufficient to describe the string given the distribution, so that the length of the description of the distribution and the size of its entropy suffices as a two-part description. The two-part description which determines the effective complexity is chosen by placing a time bound (taken from the singleton description) on the computation of the string from the distribution. The hope is that this will move deep (ie. hard to compute) information to the model part, and shallow information to the data. In a later treatment\cite{gell2004nonextensive}, Gell-Mann and Lloyd select the decisive two-part representation from the candidates based on ``the average quantities judged to be important'', which seems to be a conscious choice to make the measure subjective and context-dependant (sacrificing invariance).

G\'acs, Tromp and Vit\'anyi \cite{gacs2001algorithmic}, built on the idea of the structure function to establish a framework of \emph{algorithmic statistics}, as an attempt to found a statistical theory not on probabilistic notions, but on combinatorial ones. This echoes Kolmogorov's own hope for the theory of information:
\begin{quotation}
\noindent Information theory must precede probability theory and not be based on it. By the very essence of this discipline, the foundations of information theory must have a finite combinatorial character. \cite{kolmogorov1983combinatorial}
\end{quotation}
Algorithmic statistics studies binary strings and their possible two-part representations. In a later paper \cite{gacs2001algorithmic}, Vit\'anyi specifically studies the role of the minimal sufficient statistic as an expression of the `meaningful' information of a string (taking the name sophistication from Koppel). Like Koppel, Vit\'anyi restricts the model class to total functions, eliminating the universal model.

Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used to separate candidate two-part descriptions from those deemed to be too inefficient, is a source of instability. Small changes to this constant could shift the result sophistication by arbitrary amounts. They introduce a more elaborate criterion for the cutoff, calling this variant \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth.

Adriaans' \emph{Facticity}, introduced in 2012\cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. By constructing a very specific universal Turing machine, and taking constant terms into account, the facticity is expected to become robust against the collapse to a universal model.

\pb{Vereshchagin's strongly sufficient statistic}

\pb{Antunes' instability result}

\pb{Diagrammetje}

>>>>

\title{Three Problems for Sophistication}

@@

In this paper we describe three fundamental problems that arise with all existing proposals, showing each of them to be \emph{unsound}. Based on these results we we put forward the view that the problem is fundamental: while many nontrivial and useful structural properties of binary strings can be detected without a priori assumptions, it may be impossible to objectively quantify its sophistication.
\end{abstract}

@@

Kolmogorov complexity breaks away from the tradition in statistics and information theory by allowing the analysis of \emph{individual} binary sequences rather than studying probability distributions. This is important because for many binary strings, such as a binary encoding of a novel, or a holiday snapshot, or the fossil record, it is impossible to determine what distribution generates such objects. There is, however, a nagging discrepancy between Kolmogorov complexity and what most people whould intuitively call ``complex''. For example, a sequence of a million coin flips is exceedingly unlikely to have a Kolmogorov complexity of much less than one million bits, even though there is nothing complex about flipping a lot of coins---it's just hard work. For this reason, many scholars have defined alternative complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information in the data. This tradition started with a lecture by Kolmogorov in Talinn, Estonia in 1973, but many alternative definitions of meaningful information, or \emph{sophistication}, were proposed, see the related work section below. In this paper, we investigate three problems with these definitions, and show that they are unsatisfactory. We believe the problem to be fundamental, and provide several intuitive arguments why these attempts to define sophistication must fail.

The Kolmogorov complexity of a binary string $x$ is, roughly, the length of the shortest computer program that will output $x$. This length depends on the used programming language. However, the Invariance Theorem states that the choice between any two Turing complete languages affects the Kolmogorov complexity by at most a constant term: if $K^L(x)$ denotes Kolmogorov complexity of $x$ with respect to language $L$, then $|K^L(x)-K^M(x)|<c$ for some constant $c$ that depends on $L$ and $M$, but \emph{not on $x$}. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings. The invariance theorem allows us to fix a single reference programming language and omit it from the superscript.

A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees if it is to be interpreted as an objective property of a binary string. We are interested in a measure satisfying the following requirements:

\begin{enumerate}
\item Being an information measure, the sophistication should count the number of bits required for an effective description of the structural properties of a binary string.
\item An analogue of the Invariance Theorem should hold for sophistication: at the very least, there should be strict limits on the degree to which a change in the used programming language can affect the sophistication.
\item There should be no constant $c$ such that $S(x)\le c$ for every input sequence $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language, even for inputs of very high Kolmogorov complexity. Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that the $K(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity, and the sophistication would incorporate the amount of noise in the data just like Kolmogorov complexity does.
\end{enumerate}

For most proposed definitions of sophistication, we can prove that they fail one or more of the attributes above. Vit\'anyi's definition \cite{TODO} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only extremely deep strings (i.e., strings that require an enormous amount of processing to construct) yield a large sophistication, while intuitively it seems there must be objects that contain a lot of structure all of which can be described without much calculation. Moreover in our discussion section we provide arguments that suggest that it should be possible to generalise our negative results to show that Vit\'anyi's sophistication is not invariant.

A valid definition of sophistication must contend with the following issues. First, while all definitions are based on a two-part encoding of the data sequence, some researchers use Kolmogorov complexity as a \emph{building block} of their proposed code, while others look for the two parts of the code \emph{inside} the Kolmgorov complexity: when the definition of Kolmogorov complexity is opened up, it turns out to consist of two parts itself. In Section~\ref{sec:nickname}, we show how this second approach sets us up to suffer from to the so-called \emph{nickname problem}: the sophistication becomes highly dependent on the chosen programming language. This immediately breaks invariance, unless additional requirements are made on the language. We also show that there do exist languages (efficient enumerations of the partial recursive functions) that address this issue.

The second issue is that of \emph{overfitting}, which we consider in Section~\ref{sec:overfitting}. Overfitting is a common problem in the statistical literature, that refers to the tendency of some model selection procedures to select a very complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. For example, in a polynomial regression setting, a higher order polynomial can always provide better fit to a data set. Therefore, even if the data are in fact noisy samples from a parabola, naive model selection methods may select a much higher order polynomial that provides a worse description of the trend in the data. The same can occur for the sophistication. We somehow need to make sure that the two-part representation that determines the sophistication does not store noise in the ``structure'' part of the code. In statistics, this is typically addressed by penalising complex models, e.g. using the BIC criterion \cite{TODO}. However, such penalisations tend to break the delicate balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting (Section~\ref{sec:underfitting}) is the opposite problem, where the selected model is simple, but fails to represent part of the structure in the data. It is an especially serious problem for sophistication because the considered models for sophistication are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such interpreters can generate \emph{any} data sequence with only a short input, and if complex models are penalized to avoid overfitting, then the problem becomes to make sure that interpreters are not \emph{always} preferred for complex data. This problem is widely recognised, and is sometimes referred to as the ``collapse'' of the structure function \cite{}. The usual workaround is to restrict the set of considered models such that interpreters are no longer allowed. For example, the model class may be restricted to the set of all total recursive functions (only programs that never hang) rather than all partial recursive functions (all programs). While this patches the ``collapse'' problem to an extent, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, in the discussion in Section~\ref{sec:discussion} we argue that while two-part coding can yield useful insights into the structure of the data and allows us to identify some models as unlikely to capture all structure in the data, it is probably not possible to uniquely separate structure from noise and identify a \emph{single} model as ``best'': in general many models, of vastly different complexities, may be reasonable models of the data generating mechanism. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that sometimes the data allows for multiple very different interpretations, and that there is no such thing as sophistication.

\subsection{Related work}
The idea of separating structure from noise originated with Kolmogorov's lecture in Tallinn, Estonia in 1973 about the \emph{structure function},. This work was never published, but summarized by Cover \cite{cover1985kolmogorov}.  In this theory, the two-part description of the data consists of a program that generates a finite set, and an index of the original object in the set. The length of the program then measures the structure in the data. This approach was discussed more extensively, and generalised, by Vereshchagin, G\'acs and Vit\'anyi \cite{vereshchagin2004kolmogorov,gacs2001algorithmic}).

Koppel \cite{koppelSoph1988,koppel1991almost} also built on Kolmogorov and Cover's work to define a measure of the meaningful information in a binary string, called \emph{sophistication}, the term we adopt in this paper. He changed the class of models from finite sets to total functions and used monotone functions, instead of prefix-free Kolmogorov complexity, to be able to extend the theory to infinite strings, an approach which has been abandoned by subsequent authors. We will only consider the finite version (based on regular functions) in this paper.

In 1996 Gell-Mann and Lloyd introduced `Effective complexity' \cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (largely independently, it seems) at a notion that fits the mold of sophistication.

Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used by Koppel to separate candidate two-part descriptions from those deemed to be too inefficient, is a source of instability, a result in the spirit of this paper. They propose a different definition called \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth.

Adriaans' \emph{Facticity}, introduced in 2012\cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. Adriaans also identifies the nickname problem discussed in Section~\ref{sec:nickname}.
